---
title: "final project"
output: html_notebook
---
first we are going to load the files  

```{r}

Input = "C:/Repositories/final project/data/"
load(paste0(Input,"mat.f.coding.RData"), verbose = "TRUE")

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

Now we choose part from the body that we are going to do reserch on
```{r}
load(paste0(Input,"pheno.f.RData"), verbose = "TRUE")
ts.list = as.character(unique(pheno.f$SMTSD))
print(ts.list[53])
```
we choose the whole blood as the body part we inspect

now we want to filter the data
delete genes with low values - with 80% of expression is 0.1.

```{r}
get.raw.tissue.edata<-function(tissue.name, mat.f.coding, pheno.f){
 tiss.cols.1 = which(pheno.f$SMTSD %in% tissue.name)
 mat.1 = mat.f.coding[, tiss.cols.1]
 return(mat.1)
}
```

```{r}
i = 53
tmp.tissue = ts.list[i]
print(paste0("loading ", tmp.tissue, " edata"))
reads.src1 = get.raw.tissue.edata(tmp.tissue, mat.f.coding, pheno.f)
t.reads.src = t(reads.src1)
```

```{r}
#delete genes with low values - with 80% of expression is 0.1.
vec.1 = apply(reads.src1 , 1, function(x) length(which( x > log(0.1+1, 2) )))
row.index = which(vec.1 > (0.8*(ncol(reads.src1 ))))
# leave just rows with expression per at least 80% of the samples
src.reads = reads.src1 [row.index, ]
```
7786802 elemnts in reads.src1
4500243 elemnts after filter


```{r}
#delete genes with variance = 0
 var.data <- apply(src.reads, 1, var) #generate variance of each row - gene
 low.var.indxs = which(var.data == 0)
 if(length(low.var.indxs) > 0)
 {
 data.free = src.reads
 #now we get smaller matrix, with no genes with variance 0
 src.reads <- data.free[-low.var.indxs,]
 }
```

now we delete outliers 
The adjacency function needs WGCNA package installed


```{r}
#install.packages("BiocManager")

#BiocManager::install(c("GO.db", "preprocessCore", "impute"));
#BiocManager::install("WGCNA")
library("WGCNA")


```

```{r}
remove.outliers.with.SD<-function(t.reads.src)
{
 #remove outliers
 #cluster the samples and not the genes to find outliers
 A = adjacency(t(t.reads.src), type = "distance")
 #the connectivity of each human. -1 is to remove the diagonal, the cor to itself
 k = as.numeric(apply(A,2,sum))-1
 Z.k = scale(k) #standardized k
 thresholdZ.k = -3 #standard deviation
 outlierColor = ifelse(Z.k<thresholdZ.k,"red","black")#the red is the outlier
 my.outliers = which(outlierColor == "red")
 #printing the outlier samples
 my.outliers.samp = (rownames(t.reads.src))[my.outliers]
 print("outlier samples to remove")
 print(my.outliers.samp)

 my.good.samples = which(outlierColor == "black")
 my.good.samples.names = (rownames(t.reads.src))[my.good.samples]
 #printing the outlier samples
 #print(my.good.samples.names)
 #this is the final mat after outliers removal
 t.reads.src = t.reads.src[my.good.samples.names, ]
 return(t.reads.src)
}

```

```{r}
t.reads.src = remove.outliers.with.SD(t(src.reads))
tissue.edata = t(t.reads.src)
```

hclust try 

```{r}
sampleTree = hclust(dist(t(tissue.edata)), method = "average")
 # Plot the sample tree: Open a graphic output window of size 12 by 9 inches
 # The user should change the dimensions if the window is too large or too small.
 #sizeGrWindow(12,9)
 #pdf(file = "Plots/sampleClustering.pdf", width = 12, height = 9);
par(cex = 0.3);#change this to change the size of the text
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="",
cex.lab = 1.5,
cex.axis = 1.5, cex.main = 2)
```

quantile normalization

```{r}
#BiocManager::install("preprocessCore")
library(preprocessCore)
#rows are genes, columns are samples
quantile.normalize.raw.gtex <- function(edata.mat)
{
 norm_edata = normalize.quantiles(as.matrix(edata.mat))
 rownames(norm_edata) = rownames(edata.mat)
 colnames(norm_edata) = colnames(edata.mat)
 return(norm_edata)
}
tissue.edata.qn = quantile.normalize.raw.gtex(tissue.edata)
```
need to show box plot of the data
this is end of part 1 (lab 7 end).

